---
title: "Module3 : Getting friends with R"
author: "N. V. Schenk"
format: 
    html:
      toc: true
      self-contained: true
editor: source
execute: 
  cache: true
---
```{r, echo = F}
# just a little test
```


```{css, echo = FALSE}
.output {
max-height: 300px;
overflow-y: scroll;
}
```

```{r}
#| label: requirements
#| include: false
# load and install necessary R packages
# This is not shown in the html script.

# install necessary R packages
if(!require(dplyr)) install.packages("tidyverse")
# if(!require(cowplot)) install.packages("cowplot")

# load necessary R packages
library(tidyverse)
```


This document describes Module 4 of the course "Introduction to R", held in Summer 2023 for the [Biodiversity Exploratories](www.biodiversity-exploratories.de).

# Learning outcome

- data wrangling II
  - reshaping
- data exploration II
- reproducible workflow


:::{.callout-tip}
In order to learn coding, you need to code
:::

# Reproducible coding

Chapter 5 R4DS
Chapter 7 R4DS
maybe : Quarto : Chapter 29 R4DS

TODO : take most important of both chapters!



# Data wrangling II

Chapter 6 R4DS

- also include reshape2 package, melt and dcast
- merge
- find some common problems in data wrangling? see my notes
  - e.g. cbind and rbind cause problems
  - check R buddy intro



# Data exploration

what is missing here?
maybe add PCA?
maybe even simple stats? Or importance of defining research question?




# R code

::: {.callout-tip icon=false}

### Checkpoint 3.3

Please invent and write down 1-2 exercises including solutions about the past section (Data exploration in general, including correlation plots). You can slightly modify one of the above or the book exercises, or think about a new one.

Submit the exercises via ILIAS.

I will collect the exercises of everyone during the course, and provide them to you for the last module.

:::

::: {.panel-tabset}
## Exercise

## Solution

:::

