# The EML R package

## Aim
create machine-readable metadata, using the Ecological Metadata Language.
This package creates an .XML format metadata, which is machine-readable.

For human-readability, it is convenient to convert the dataset to .HTML, e.g. with the online tool : https://codebeautify.org/xml-to-html-converter


## Material
- R package vignette : https://cran.r-project.org/web/packages/EML/EML.pdf
- website with background information, recommended/ writte by developers : https://eml.ecoinformatics.org/
- GitHub of package (with most recent information) : https://github.com/ropensci/EML
- Tutorial to follow (EML Tutorial): https://nceas.github.io/oss-lessons/ecological-metadata/ecological-metadata.html

- minimal tutorial on the `xml2` package : https://www.rstudio.com/blog/xml2/

The citation abbreviation used in this document is given in brackets.

## INSTALL PACKAGE
*Note* on Ubuntu OS : non-zero exit status of the EML package installation : requires you to install the libjq-dev package (not within R, but within your operating system)
```{r}
# install.packages("EML")
library(EML)
```

## Minimal example 
Taken from the tutorial linked above (EML Tutorial).
```{r}
me <- list(individualName = list(givenName = "Carl", surName = "Boettiger"))
my_eml <- list(dataset = list(
              title = "A Minimal Valid EML Dataset",
              creator = me,
              contact = me)
            )
write_eml(my_eml, "ex.xml")
eml_validate("ex.xml")
```


## read, navigate and visualise
A xml file is not very easy to understand for humans (it's machine-readable). But in R, you can e.g. convert it to a list in order to read it more easily.
```{r}
# install.packages("xml2")
library(xml2)
f <- read_xml("ex.xml")
as_list(f) # see that not all fields are filled.
```
