---
title: "Reshape datasets"
author: "N. Schenk"
date: "2022-11-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Reshaping datasets

## Requirements
```{r}
library(ggplot2)
library(reshape2)
```


## Build example dataset
Plant species abundance, samples in 100 plots for 4 species.
```{r}
# create an example dataset "plantspecies"
plantspecies <- data.frame(plotID = as.factor(seq(1, 100, 1)),
                           landuse_gradient = sample(seq(1, 3, 0.1), 100, replace = T),
                           plottaria_exampulus = 2*landuse_gradient + 12 + rnorm(100, mean = 12/3, sd = 12/0.5),
                           graphica_demonstrandi = 3*landuse_gradient + 8 + rnorm(100, mean = 8/2, sd = 8/1),
                           arrclubus_cursus = 2.5*landuse_gradient + 25 + rnorm(100, mean = 25/2, sd = 25/2),
                           exemplaria_plotii = 3*landuse_gradient + 9 + rnorm(100, mean = 9/2, sd = 9/0.8))
head(plantspecies)
```

## Transform from wide to long format

The dataset is in wide format : every plant species has its own column. This is very convenient to plot results for individual plant species (see example below), but not for plotting overviews like in a boxplot of abundance per plantspecies.
```{r}
# only possible to plot 1 plant species at a time
# R base plots
plot(plantspecies$landuse_gradient, plantspecies$plottaria_exampulus)
plot(plantspecies$landuse_gradient, plantspecies$graphica_demonstrandi)
# GGplot example
ggplot(plantspecies, aes(x = landuse_gradient, y = plottaria_exampulus)) +
  geom_point()
```

Transform to long format
```{r}
# bring to ggplot format, using the package reshape2
plantspecies_long <- melt(plantspecies, 
          id.vars = c("plotID", "landuse_gradient"),
          measure.vars = c("plottaria_exampulus", "graphica_demonstrandi", "arrclubus_cursus", "exemplaria_plotii"),
          value.name = "Abundance",
          variable.name = "Species")
# minimal example
melt(plantspecies, id.vars = c("plotID", "landuse_gradient"))
# all the other parameters are guessed from the input = the function knows itself what to do
# if you just tell it which columns need to stay.

# Plot dataset
ggplot(plantspecies_long, aes(x = Species, y = Abundance)) +
  geom_boxplot() +
  geom_point()
```
